Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-03 paper/manuscript.tex
Table 1 legend

\subsection*{Two Env homologs from clade A transmitted-founder viruses}
The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by our group~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The sequence alignment of clade A \textit{env} coding sequences in FASTA format is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

\subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain of HIV~\citep{haddox2016experimental}.
In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably identify differences that exceed the magnitude of the experimental noise.

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce the sequencing error rate.
We also deep sequenced identically handled wildtype controls to estimate error and mutation rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking the library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries mutagenized 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to make comparisons to the earlier inferior LAI data set. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across both Envs, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
This fact is an initial hint that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

\subsection*{Re-scaling the experimental measurements to optimally describe HIV evolution in nature}
Next, we sought to compare Env's amino-acid preferences between BG505 and BF520.
We also sought to compare these homologs to LAI~\cite{haddox2016experimental}.
However, a concern is that differences we see between homologs may be due to experimental differences, rather than true biological differences.
For instance, the LAI experiments involved passaging the viral libraries for a substantially longer amount of time in cell culture, which may have imposed increased selection.
Additionally, even though we conducted the BG505 and BF520 experiments using very similar protocols, the experiments seemed to have exerted different levels of purifying selection, as indicated by differential purging of stop codons (Fig \ref{fig:ch3_mutfreqs}A).
The strength of selection in our experiments may also differ from the strength of selection in nature.
Ideally, however, we would like to compare the preferences after rescaling them to reflect the intensity of selection in a natural setting.

@@

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Phylogenetic models that incorporate Env's preferences indicate that selection was less stringent in the lab than in nature.}
\tabledata{Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

@@

In contrast to BG505 and BF520, for both alignments, the ExpCMs inferred a stringency parameter of $\sim$1 for LAI (Table \ref{tab:ch3_phydms}).
Thus, both the stringency parameter and the overall model performance was lower for LAI than for the other homologs.
One explanation for these differences is biological.
Perhaps the preferences for LAI -- a lab-adapted virus isolated from chronic infection -- are simply less representative of the dominant selective pressures that shape Env's evolution in nature.
An alternative explanation is experimental.
It is possible that if we repeated the LAI deep mutational scan using the same method we used for BG505 and BF520, our estimates would change.
We are currently repeating these experiments.
Thus, for the remainder of this chapter, I will only focus on comparing the preferences for BG505 and BF520.

>>>>

\subsection*{Two Env homologs from clade A transmitted-founder viruses}
The viruses most relevant to HIV's long-term evolution are those which are transmitted from human-to-human.
However, the only prior work that has measured how all Env amino-acid mutations affect HIV replication is a study by some of us~\citep{haddox2016experimental} that used a late-stage lab-passaged CXCR4-tropic virus~\citep[LAI;][]{peden1991changes}.
The properties of Env can vary substantially between such late-stage viruses and the transmitted-founder viruses relevant to HIV's long-term evolution~\citep{sagar2006human,wilen2011phenotypic,parrish2013phenotypic}.

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

\subsection*{Deep mutational scanning to measure the effects of all amino-acid mutations to each Env on HIV replication}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV replication in cell culture, and applied this strategy to the LAI strain~\citep{haddox2016experimental}.
In addition to using a late-stage lab-adapted HIV strain, that study had another shortcoming: the measurements were quite noisy.
Specifically, the Pearson correlation coefficients between measurements of mutational effects from independent biological replicates were in the range of $R = 0.45$ to $0.50$~\citep{haddox2016experimental}.
This noise is a problem when trying to identify differences in mutational effects between Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise.

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries mutagenized 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, we focus on comparisons between BG505 and BF520, and do not attempt to compare to the earlier inferior LAI data set. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across both Envs, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
This fact is an initial hint that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

\subsection*{Amino-acid preferences of the BG505 and BF520 Envs and their relationship to HIV evolution in nature}
Next, we sought to compare Env's amino-acid preferences between BG505 and BF520.
We also sought to compare these homologs to LAI~\cite{haddox2016experimental}.
However, a concern is that differences we see between homologs may be due to experimental differences, rather than true biological differences.
For instance, the LAI experiments involved passaging the viral libraries for a substantially longer amount of time in cell culture, which may have imposed increased selection.
Additionally, even though we conducted the BG505 and BF520 experiments using very similar protocols, the experiments seemed to have exerted different levels of purifying selection, as indicated by differential purging of stop codons (Fig \ref{fig:ch3_mutfreqs}A).
The strength of selection in our experiments may also differ from the strength of selection in nature.
Ideally, however, we would like to compare the preferences after rescaling them to reflect the intensity of selection in a natural setting.

@@

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing either deep mutational scanning dataset describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1, indicating that natural evolution prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
All models draw $\omega$ (the dN/dS ratio for the Goldman-Yang model, and the relative dN/dS ratio after accounting for the measured constraints for the ExpCM) from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results of analyses of the same tree with additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

@@
